---
title: "ISA 616 Workflow"
author: "Andrew Bubnar"
date: "10/1/2020"
output: 
  html_document:
    theme: cerulean
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

# Optimizing Capital Bikeshare Product Distribution
##### An Analysis by Andrew Bubnar

## Overview
I believe that there is room to optimize the distribution process of Capital Bikeshare through predictive modeling. Capital Bikeshare is a bikesharing company, owned by Lyft Inc., that allows registered and casual users to rent public bikes in major cities around the world. Currently, bikes are deployed uniformly over some time dimensions. This creates a lapse in efficiency, that is, bikes are not always available when they need to be, and some bikes are sitting idle, producing no revenue. If we can predict the number of bikes that will be in use at any given time, we will maximize use, and thus profitability. 

The purpose of this analysis is, firstly, to deduce what factors impact the outcome of riders. The data that will be considered is primarily related to time, weather, and expected traffic. After concluding the impacts of each variable, suggestions to Capital Bikeshare's Logistics Controls team will be made (and passed along to distributors), to allow for a more satisfied customer base and increased profitabilty.

```{r, include = FALSE }
# Initialize all Packages that will be used
if(require(pacman)==FALSE)
  install.packages("pacman")

pacman::p_load(tidyverse, hexbin, dplyr, DT, ggplot2, caret, corrplot, knitr)

```

```{r }

```


## Data Selection, Description and Summary

```{r }

```


#### Data Sourcing

```{r }

```


#### Data Head & Codebook

```{r }

```


#### Data Structure

```{r }

```


#### Data Missingness

```{r }

```


#### Two Options for Analysis

```{r }

```


## Data Processing

```{r }

```


## Modeling



```{r }


```

## Final Model and Justification


```{r }

```

## Interpretation and Conclusion


```{r }

```

## Source






